Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-12 paper.tex
Change title
Updates

\title{Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2}
\author[1]{Shing H. Zhan}
\author[2,5]{Katherine Eaton}
\author[3,5]{Anastasia Ignatieva}
\author[1,5]{Benjamin Jeffery}
\author[1,5]{Duncan S. Palmer}
\author[1,5]{Yan Wong}
\author[2]{Carmen L. Murall}
\author[4]{Sarah P. Otto}
\author[1,6]{Jerome Kelleher}
\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
\affil[5]{Preliminary alphabetic ordering}
\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.

@@

These sc2ts backbone phylogenies do contain small amounts of recombination

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the leftmost local tree in the Wide ARG
(sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees are pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

>>>>

\title{Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
\author[1]{Shing H. Zhan}
\author[2,5]{Katherine Eaton}
\author[3,5]{Anastasia Ignatieva}
\author[1,5]{Benjamin Jeffery}
\author[1,5]{Duncan S. Palmer}
\author[1,5]{Yan Wong}
\author[2]{Carmen L. Murall}
\author[4]{Sarah P. Otto}
\author[1,6]{Jerome Kelleher}
\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
\affil[5]{Preliminary alphabetic ordering}
\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the leftmost local tree in the Wide ARG
(sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees are pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

@@

Our sc2ts backbone phylogenies do contain small amounts of recombination:
7 recombination nodes in the backbone of the wide ARG and 8 in the backbone
of the long ARG. However, the backbones remain extremely tree-like, as these
recombination events only involve minor and local topological rearrangements:
for instance recombination between two sibling lineages, or even recombination
within a single lineage (in which the parent lineages above a recombination event
coalesce with each other before coalescing with any other lineages in
the phylogeny
